Phylodynamic inference of contact network parameters through approximate Bayesian computation

### 2016-06-14 doc/thesis/objective.tex
Editing and proofreading intro

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population~\autocite{hamer1906milroy,
kermack1927contribution}, or a population divided into a small number of
homogeneously mixed groups~\autocite{rushton1955deterministic}. This
assumption, also called \defn{mass
action}~\autocite{heesterbeek2000mathematical}, or \defn{panmixia}, implies
that any two individuals in the same compartment are equally likely to come
into contact making transmission possible at some predefined rate. Although
this provides a reasonable approximation in many
cases~\autocite{anderson1992infectious}, the error introduced by assuming a
panmictic population can be substantial when significant contact heterogeneity
exists in the underlying
population~\autocite{bansal2007individual,barthelemy2005dynamical,keeling2005networks}.
Contact network models provide an alternative to compartmental models which do
not require the assumption of panmixia. In addition to more accurate
predictions, the parameters of the networks themselves may be of interest from
a public health perspective. For example, certain vaccination strategies may be
more or less effective in curtailing an epidemic depending on the underlying
network's degree distribution~\autocite{peng2013vaccination, ma2013importance}.
Phylodynamic methods, {\color{blue}\uline{which link viruses' evolutionary and
epidemiological dynamics}}, have been used to fit many different types of models
to phylogenetic data~\autocite{pybus2009evolutionary,volz2013viral}. However,
these models generally assume a panmictic population. The primary objective of
this work is \emph{to develop a method to fit contact network models,
{\color{blue}\uline{and thereby relax the assumption of homogeneous mixing}},
in a phylodynamic framework.}

@@

{\color{blue}\uline{In this work, we take a Bayesian approach: our goal is to
estimate the posterior distribution on model parameters given our data,}
\[
    \Pr(\theta \mid T) = \frac{\Pr(T \mid \theta) \Pr(\theta)}{\Pr(T)},
\]
\uline{where $\Pr(T \mid \theta)$ is the likelihood, $\Pr(\theta)$ is the prior,
and $\Pr(T)$ is the marginal probability of $T$ which acts as a normalizing
constant on the posterior (see} \cref{chp:prelim} \uline{for a review of
mathematical modeling and Bayesian inference, including definitions of these
concepts). As we shall show} (see \cref{subsec:just}), \uline{estimating this
distribution presents computational challenges beyond those usually encountered
in Bayesian inference. Both the likelihood $\Pr(T \mid \theta)$ and the
normalizing constant $\Pr(T)$ seem to be intractable, which rules out the use
of most common \acrlong{ML} and Bayesian methods. }}

{\color{red}\sout{Calculating the likelihood of the parameters of a contact
network model seems likely to be an intractable problem. We have not proven
this is the case, but some intuition can be provided by examining the process
involved in the likelihood calculation. Consider a contact network model with
parameters $\theta$ and an estimated transmission tree $T$ with $n$ tips. In
general, we do not know the labels of the internal nodes of $T$, only the
labels of its tips. To fit this model using likelihood-based methods, we must
calculate the likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$
be the set of all possible contact networks, and $\Nu$ be the set of all
possible labellings of the internal nodes of $T$. We can write the likelihood
as}
\begin{align}
\begin{split}
  \label{eq:netlik}
  \Pr(T \mid \theta)
    &= \sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta) \\
    &= \sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),
\end{split}
\end{align}
\sout{the last equality following from the fact that $T$ and $\nu$ depend only
on $G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
may individually be straightforward to calculate, the number of possible
directed graphs on $N$ nodes is $2^{N(N-1)}$~\autocite{harary2014graphical},
larger if the nodes and edges in the graph may have different labels or
attributes. Hence, the number of terms in the sum is at least exponential in
$n$, as there must be at least $n$ nodes in the network. In addition,
}\cref{eq:netlik} \sout{assumes that $T$ is complete, meaning that all infected
individuals were sampled. This is rarely the case in practice - most often, we
only have access to a subset of the infected individuals. In this case, the
likelihood calculation becomes even more complex, because we must also sum over
all possible complete trees.}}

{\color{red}\sout{Depending on the network model studied, it is possible that}
\cref{eq:netlik} \sout{could be simplified into a tractable expression.}}
An alternative to likelihood-based methods, which could be applied to any
network model, is provided by \gls{ABC}~\autocite{rubin1984bayesianly,
tavare1997inferring, fu1997estimating, beaumont2002approximate}. All of the
ingredients required to apply \gls{ABC} to this problem are readily available.
Simulating networks is straightforward under a variety of models. Epidemics on
those networks, and the corresponding transmission trees, can also be easily
simulated. As mentioned above, contact networks can profoundly affect
transmission tree shape. Those shapes can be compared using a highly
informative similarity measure called the ``tree
kernel''~\autocite{poon2013mapping}; {\color{blue}\uline{similar kernel
functions have been demonstrated to work well as distance functions in
\gls{ABC}~\autocite{park2015k2}}}. \Gls{ABC} can be implemented with several
algorithms, but \gls{SMC} has advantages over others,
{\color{blue}\uline{including improved accuracy in low-density regions and
parallelizability}}~\autocite{mckinley2009inference}. A recently-developed
adaptive algorithm requiring minimal tuning on the part of the user makes
\gls{SMC} an even more attractive approach~\autocite{del2012adaptive}. In
summary, our method to infer contact network parameters will combine the
following: stochastic simulation of epidemics on networks, the tree kernel, and
adaptive \gls{ABC}-\gls{SMC}. {\color{blue}\uline{ Our method will expand on
the framework developed by~\autocite{poon2015phylodynamic}, who combined
\gls{ABC} with the tree kernel to infer parameters of population genetic models
from viral phylogenies using \gls{MCMC}. }}

@@

{\color{blue}\uline{
The remainder of this background chapter is organized in four sections. The
first section introduces phylogenies and transmission trees, which are the
input data from which our method aims to make statistical inferences. This
section also introduces phylodynamics, a family of methods that, like ours, aim
to infer epidemiological parameters from evolutionary data. The second section
focuses on contact networks and network models, whose parameters we are
attempting to infer. The relationship between contact networks and transmission
trees is also discussed. The third and fourth sections introduce \gls{SMC}
and \gls{ABC} respectively, which are the two algorithmic components of the
method we will implement. In particular, \gls{ABC} refers to the general
approach of using simulations to replace likelihood calculations in a Bayesian
setting, while \gls{SMC} is a particular algorithm which can be used to
implement \gls{ABC}. }}

>>>>

The spread of a disease is most often modelled by assuming either a
homogeneously mixed population~\autocite{hamer1906milroy,
kermack1927contribution}, or a population divided into a small number of
homogeneously mixed groups~\autocite{rushton1955deterministic}. This
assumption, also called \defn{mass
action}~\autocite{heesterbeek2000mathematical}, or \defn{panmixia}, implies
that any two individuals in the same compartment are equally likely to come
into contact making transmission possible at some predefined rate. Although
this provides a reasonable approximation in many
cases~\autocite{anderson1992infectious}, the error introduced by assuming a
panmictic population can be substantial when significant contact heterogeneity
exists in the underlying
population~\autocite{bansal2007individual,barthelemy2005dynamical,keeling2005networks}.
Contact network models provide an alternative to compartmental models which do
not require the assumption of panmixia. In addition to more accurate
predictions, the parameters of the networks themselves may be of interest from
a public health perspective. For example, certain vaccination strategies may be
more or less effective in curtailing an epidemic depending on the underlying
network's degree distribution~\autocite{peng2013vaccination, ma2013importance}.
Phylodynamic methods, \add{which link viruses' evolutionary and epidemiological
dynamics}, have been used to fit many different types of models to phylogenetic
data~\autocite{pybus2009evolutionary,volz2013viral}. However, these models
generally assume a panmictic population. The primary objective of this work is
\emph{to develop a method to fit contact network models, \add{and thereby relax
the assumption of homogeneous mixing,} in a phylodynamic framework.}

@@

\add{In this work, we take a Bayesian approach: our goal is to
estimate the posterior distribution on model parameters given our data,}
\[
  \madd{\post(\theta \mid T) = \frac{\lik(T \mid \theta) \prior(\theta)}
    {\int_\Theta \lik(T \mid \theta) \prior(\theta) \d \theta}},
\]
\add{where $\lik(T \mid \theta)$ is the likelihood of the parameters given $T$,
$\prior(\theta)$ is the prior on $\theta$, and $\Theta$ is the space of
possible model parameters. The denominator on the right-hand side is the
marginal probability of $T$ which acts as a normalizing constant on the
posterior (see \cref{chp:prelim} for a review of mathematical modeling and
Bayesian inference, including definitions of these concepts). As we shall show
(\cref{subsec:just}), estimating this distribution presents computational
challenges beyond those usually encountered in Bayesian inference. Both the
likelihood $\lik(T \mid \theta)$ and the normalizing constant seem to be
intractable, which rules out the use of most common \acrlong{ML} and Bayesian
methods.}

\del{Calculating the likelihood of the parameters of a contact network model
seems likely to be an intractable problem. We have not proven this is the case,
but some intuition can be provided by examining the process involved in the
likelihood calculation. Consider a contact network model with parameters
$\theta$ and an estimated transmission tree $T$ with $n$ tips. In general, we
do not know the labels of the internal nodes of $T$, only the labels of its
tips. To fit this model using likelihood-based methods, we must calculate the
likelihood of $\theta$, that is, $\Pr(T \mid \theta)$. Let $\G$ be the set of
all possible contact networks, and $\Nu$ be the set of all possible labellings
of the internal nodes of $T$. We can write the likelihood as}
\begin{align}
\begin{split}
  \label{eq:netlik}
  \mdel{\Pr(T \mid \theta)}
    &\mdel{=\sum_{\nu \in \Nu} \Pr(T, \nu \mid \theta)} \\
    &\mdel{=\sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G, \theta) \Pr(G \mid \theta)} \\
    &\mdel{=\sum_{G \in \G} \sum_{\nu \in \Nu} \Pr(T, \nu \mid G) \Pr(G \mid \theta),}
\end{split}
\end{align}
\del{the last equality following from the fact that $T$ and $\nu$ depend only
on $G$, not on $\theta$. Although $\Pr(T, \nu \mid G)$ and $\Pr(G \mid \theta)$
may individually be straightforward to calculate, the number of possible
directed graphs on $N$ nodes is $2^{N(N-1)}$~\autocite{harary2014graphical},
larger if the nodes and edges in the graph may have different labels or
attributes. Hence, the number of terms in the sum is at least exponential in
$n$, as there must be at least $n$ nodes in the network. In addition,
\cref{eq:netlik} assumes that $T$ is complete, meaning that all infected
individuals were sampled. This is rarely the case in practice - most often, we
only have access to a subset of the infected individuals. In this case, the
likelihood calculation becomes even more complex, because we must also sum over
all possible complete trees.}

\del{Depending on the network model studied, it is possible that
\cref{eq:netlik} could be simplified into a tractable expression.} An
alternative to likelihood-based methods, which could be applied to any network
model, is provided by \gls{ABC}~\autocite{rubin1984bayesianly,
tavare1997inferring, fu1997estimating, beaumont2002approximate}. All of the
ingredients required to apply \gls{ABC} to this problem are readily available.
Simulating networks is straightforward under a variety of models. Epidemics on
those networks, and the corresponding transmission trees, can also be easily
simulated. As mentioned above, contact networks can profoundly affect
transmission tree shape. Those shapes can be compared using a highly
informative similarity measure called the ``tree
kernel''~\autocite{poon2013mapping}; \add{similar kernel functions have been
demonstrated to work well as distance functions in
\gls{ABC}~\autocite{park2015k2}}. \Gls{ABC} can be implemented with several
algorithms, but \gls{SMC} has advantages over others,
\add{including improved accuracy in low-density regions and
parallelizability}~\autocite{mckinley2009inference}. A recently-developed
adaptive algorithm requiring minimal tuning on the part of the user makes
\gls{SMC} an even more attractive approach~\autocite{del2012adaptive}. In
summary, our method to infer contact network parameters will combine the
following: stochastic simulation of epidemics on networks, the tree kernel, and
adaptive \gls{ABC}-\gls{SMC}. \add{Our method will expand on the framework
developed by~\autocite{poon2015phylodynamic}, who combined \gls{ABC} with the
tree kernel to infer parameters of population genetic models from viral
phylogenies using \gls{MCMC}.}

@@

\add{The remainder of this background chapter is organized in four sections.
The first section introduces phylogenies and transmission trees, which are the
input data from which our method aims to make statistical inferences. This
section also introduces phylodynamics, a family of methods that, like ours, aim
to infer epidemiological parameters from evolutionary data. The second section
focuses on contact networks and network models, whose parameters we are
attempting to infer. The relationship between contact networks and transmission
trees is also discussed. The third and fourth sections introduce \gls{SMC}
and \gls{ABC} respectively, which are the two algorithmic components of the
method we will implement. In particular, \gls{ABC} refers to the general
approach of using simulations to replace likelihood calculations in a Bayesian
setting, while \gls{SMC} is a particular algorithm which can be used to
implement \gls{ABC}.}

### 2016-06-14 doc/thesis/preamble.tex
Editing and proofreading intro

\documentclass[11pt]{report}
\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage[top=1in, bottom=1in, inner=1.25in, outer=1in]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
\usepackage{nicefrac}
\usepackage{amsmath, amssymb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{multicol}
\usepackage{multirow}
\usepackage[hidelinks]{hyperref}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{glossary-mcols}
\usepackage{cleveref}
\usepackage{array}
\usepackage{textcomp}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{fancyhdr}
\usepackage[dvipsnames]{xcolor}
\usepackage[titletoc]{appendix}
\usepackage{tocloft}
\usepackage[normalem]{ulem}

@@

\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
\newcommand{\sett}[1]{\{#1\}}
\renewcommand{\emptyset}{\varnothing}
\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\star}[1]{#1^*}
\renewcommand{\d}{\mathrm{d}\,}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\maps}{:}

@@

\newfloat{data}{hb}{lod}
\floatname{data}{Data}
\crefname{data}{data}{data}

@@

\newglossaryentry{alpha}{
  name=\ensuremath{\alpha},
  description=preferential attachment power parameter in Barab\'asi-Albert networks
}
\newglossaryentry{m}{
  name=\ensuremath{m},
  description=number of edges added per vertex when constructing a Barab\'asi-Albert network
}
\newglossaryentry{N}{
  name=\ensuremath{N},
  description=total number of nodes in a contact network
}
\newglossaryentry{I}{
  name=\ensuremath{I},
  description=number of infected nodes in a contact network at the time of
  transmission tree sampling
}
\newglossaryentry{gamma}{
  name=$\gamma$,
  description=exponent of power-law degree distribution in scale-free networks
}
\newglossaryentry{p}{
  name=\ensuremath{p},
  description=edge density parameter in Erd\H{o}s-R\'enyi networks
}
\newglossaryentry{lambda}{
  name=\ensuremath{\lambda},
  description=decay factor meta-parameter for tree kernel
}
\newglossaryentry{sigma}{
  name=\ensuremath{\sigma},
  description=radial basis function variance meta-parameter for tree kernel
}
\newglossaryentry{n}{
  name=\ensuremath{n},
  description=number of particles used for ABC-SMC
}
\newglossaryentry{M}{
  name=\ensuremath{M},
  description=number of simulated datasets per particle in ABC-SMC
}
\newglossaryentry{alphaess}{
  name=\ensuremath{\alpha_{\ESS}},
  description=per-iteration rate of expected sample size decay in adaptive ABC-SMC
}

>>>>

\documentclass[11pt]{report}
\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage[top=1in, bottom=1in, inner=1.25in, outer=1in]{geometry}
\usepackage{setspace}
\usepackage[backend=biber, style=numeric-comp, sorting=none]{biblatex}
\usepackage{nicefrac}
\usepackage{amsmath, amssymb}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{multicol}
\usepackage{multirow}
\usepackage[hidelinks]{hyperref}
\usepackage[acronym,nonumberlist]{glossaries}
\usepackage{glossary-mcols}
\usepackage{cleveref}
\usepackage{array}
\usepackage{textcomp}
\usepackage{float}
\usepackage{pdflscape}
\usepackage{fancyhdr}
\usepackage[dvipsnames]{xcolor}
\usepackage[titletoc]{appendix}
\usepackage{tocloft}
\usepackage[normalem]{ulem}
\usepackage{calc}

@@

\newcommand{\set}[1]{\left\lbrace#1\right\rbrace}
\newcommand{\sett}[1]{\{#1\}}
\renewcommand{\emptyset}{\varnothing}
\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\star}[1]{#1^*}
\renewcommand{\d}{\mathrm{d}\,}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\maps}{:}
\newcommand{\vx}{\vec{x}}

\newcommand{\prior}{\pi}
\newcommand{\lik}{f}
\newcommand{\post}{\pi}

\let\mref\cref
\let\Mref\Cref
\let\mtextcite\textcite
\let\mautocite\autocite
\renewcommand{\cref}[1]{\mbox{\mref{#1}}}
\renewcommand{\Cref}[1]{\mbox{\Mref{#1}}}
\renewcommand{\textcite}[1]{\mbox{\mtextcite{#1}}}
\renewcommand{\autocite}[1]{\mbox{\mautocite{#1}}}
\newcommand{\add}[1]{\color{blue}\uline{#1} \color{black}}
\newcommand{\madd}[1]{\color{blue}\underline{#1} \color{black}}
\newcommand{\del}[1]{\color{red}\sout{#1} \color{black}}
% http://tex.stackexchange.com/a/20612/60841
\newsavebox\CBox
\newcommand\mdel[1]{%
  \color{red}%
  \ifmmode\sbox\CBox{$#1$}\else\sbox\CBox{#1}\fi%
  \makebox[0pt][l]{\usebox\CBox}%
  \rule[0.5\ht\CBox-0.25pt]{\wd\CBox}{0.5pt}
  \color{black}%
}
%TC:macro \del [ignore]
%TC:macro \mdel [ignore]

@@

@@

\newglossaryentry{alpha}{
  name=\ensuremath{\alpha},
  description=preferential attachment power parameter in Barab\'asi-Albert networks
}
\newglossaryentry{m}{
  name=\ensuremath{m},
  description=number of edges added per vertex when constructing a Barab\'asi-Albert network
}
\newglossaryentry{N}{
  name=\ensuremath{N},
  description=total number of nodes in a contact network
}
\newglossaryentry{I}{
  name=\ensuremath{I},
  description=number of infected nodes in a contact network at the time of
  transmission tree sampling
}
\newglossaryentry{gamma}{
  name=$\gamma$,
  description=exponent of power-law degree distribution in scale-free networks
}
\newglossaryentry{p}{
  name=\ensuremath{p},
  description=edge density parameter in Erd\H{o}s-R\'enyi networks
}
\newglossaryentry{lambda}{
  name=\ensuremath{\lambda},
  description=decay factor meta-parameter for tree kernel
}
\newglossaryentry{sigma}{
  name=\ensuremath{\sigma},
  description=radial basis function variance meta-parameter for tree kernel
}
\newglossaryentry{n}{
  name=\ensuremath{n},
  description=number of particles used for sequential Monte Carlo
}
\newglossaryentry{M}{
  name=\ensuremath{M},
  description=number of simulated datasets per particle in ABC-SMC
}
\newglossaryentry{alphaess}{
  name=\ensuremath{\alpha_{\ESS}},
  description=per-iteration rate of expected sample size decay in adaptive ABC-SMC
}
\newglossaryentry{Theta}{
  name=\ensuremath{\Theta},
  description=parameter space for a given mathematical model
}

### 2016-06-14 doc/thesis/preface.tex
Editing and proofreading intro

A version of chapter 2 has been submitted for publication with the title
``Reconstructing network parameters from viral phylogenies.'' An oral
presentation entitled ``Phylodynamic inference of contact network parameters
with kernel-ABC'' was given based on chapter 2 to the 23rd HIV Dynamics and
Evolution meeting on April 25, 2016, in Woods Hole, Massachusetts, USA (the
presentation was delivered remotely). 
%A poster based on chapter 2 entitled
%``Likelihood-free estimation of contact network parameters from viral
%phylogenies'' was presented at the Intelligent Systems for Molecular Biology
%meeting in Orlando, Florida, USA.

Use of the BC data is in accordance with an ethics application that was
reviewed and approved by the UBC/Providence Health Care Research Ethics Board
(H07-02559).

>>>>

A version of chapter 2 has been submitted for publication with the title
``Reconstructing network parameters from viral phylogenies.'' An oral
presentation entitled ``Phylodynamic inference of contact network parameters
with kernel-ABC'' was given based on chapter 2 to the 23rd HIV Dynamics and
Evolution meeting on April 25, 2016, in Woods Hole, Massachusetts, USA (the
presentation was delivered remotely). A poster based on chapter 2 entitled
``Likelihood-free estimation of contact network parameters from viral
phylogenies'' is scheduled for presentation at the Intelligent Systems for
Molecular Biology meeting on July 8, 2016, in Orlando, Florida, USA.

Use of the BC data is in accordance with an ethics application that was
reviewed and approved by the UBC/Providence Health Care Research Ethics Board
(H07-02559). Rosemary M. McCloskey completed the Tri-Council Policy Statement:
Ethical Conduct for Research Involving Humans Course on Research Ethics (TCPS
2: CORE) tutorial on June 8, 2016.

\add{Source code for the \software{netabc} program is freely available at
\url{https://github.com/rmcclosk/netabc} under the GPL-3 license. Scripts to
run all computational experiments, as well as the source code for this thesis,
are available at \url{https://github.com/rmcclosk/thesis}.}
